---
title: "Descriptive Statistics"
subtitle: "Appendices"
institute: "Ethan Marzban (for DS Collab)"
author: "October 12, 2023"
footer: "[dscollab.github.io](https://dscollab.github.io)"
logo: "main_logo2.png"
format: 
  revealjs:
    theme: slides.scss
    multiplex: true
    transition: fade
    slide-number: true
    incremental: true 
    chalkboard: true
    html-math-method: mathjax
editor: visual
execute:
  freeze: auto
---

# Appendix 1: Mean vs. Median

## Robustness

-   During the workshop, I mentioned that the median is a more *robust* description of central tendency than the mean.

    -   I briefly mentioned that this means the median is less affected by outliers than the mean.

-   To illustrate this a little more explicitly, lets consider a set of numbers $$ S = \{1, 2, 3, 4\} $$

-   We will then add various numbers to this set, and see how this affects both the mean and the median.

## Illustration

-   First, note that $\overline{s} = 2.5$ and $\mathrm{median}(S) = 2.5$.

-   Now, suppose we construct a new set $S'$ defined to the the set $S$ plus the additional value $s_5 = 5$.

    -   That is, $S' = \{1, 2, 3, 4, 5\}$.

-   The mean of $S'$ is 3, and the median of $S'$ is also 3.

-   Hence, the addition of the point $s_5$ caused both the mean and median to increase by 0.5 units.

## Illustration

-   Now, suppose instead of adding the number $s_5 = 5$ we instead added the number $s_5 = 6$.

    -   That is, $S' = \{1, 2, 3, 4, 6\}$.

-   Now, the mean of $S'$ is 3.2, whereas the median of $S'$ is 3.

-   So, the addition of $s_5$ has caused the mean to increase by $0.7$ but caused the median to still only increase by 0.5 units.

## Illustration

-   If instead of adding 6 we had added 7 to obtain $S' = \{1, 2, 3, 4, 7\}$ the mean would be $3.4$ whereas the median would still be $3$.

-   We can imagine continuing this line of thinking for various values of $s_5$.

::: fragment
| Set                 | Mean  | Median |
|---------------------|-------|--------|
| $\{1, 2, 3, 4\}$    | $2.5$ | $2.5$  |
| $\{1, 2, 3, 4, 5\}$ | $3$   | $3$    |
| $\{1, 2, 3, 4, 6\}$ | $3.2$ | $3$    |
| $\{1, 2, 3, 4, 7\}$ | $3.4$ | $3$    |
:::

## Illustration

-   We can rephrase our findings to be in therms of the *change* in mean and median of $S'$ (as compared to $S$):

::: fragment
| $s_5$ | Change in Mean | Change in Median |
|-------|----------------|------------------|
| $5$   | $0.5$          | $0.5$            |
| $6$   | $0.7$          | $1$              |
| $7$   | $0.9$          | $1$              |
| $8$   | $1.1$          | $1$              |
:::

-   So, the larger $s_5$ is, the larger the change in the mean. However, the change in the median remains relatively fixed, regardless of how large $s_5$ is!

## Illustration

-   This is, effectively, the basic idea of robustness.

-   The mean changes significantly based on which point(s) we add to our data, whereas the median does not.

-   The notion of robustness isn't only tied to measures of central tendency; for example, the IQR turns out to be a more robust description of spread than the variance/standard deviation.

## Connections

-   The notion of robustness has far-reaching consequences. For example, when providing summary measures of variables with heavy-tails, the median is more often used for the mean.

-   As a concrete example, consider `income`: the distribution of incomes has a very large right-tail as there are a few individuals with incredibly high amounts of wealth.

    -   Consequently, most analysists report/consider the *median* income of an area, as opposed to the *mean* income.

# Linearity vs. Nonlinearity

## Linearity of Functions

-   Let's start by talking about linearity in the context of functions.

-   Consider two functions `f()` and `g()`, plotted below:

::: fragment
```{r, fig.height = 4}
library(tidyverse)
library(ggthemes)
library(gridExtra)

f <- function(x){return(-1.5*x)}
p1 <- data.frame(x = seq(0, 4, by = 0.1)) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = f, col = "blue", linewidth = 1) +
  theme_economist_white(base_size = 24) +
  ggtitle("f(x)") +
  theme(panel.background = element_rect("white"),
        plot.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 16,
                                    margin = margin(
                                      t = 0, 
                                      r = 10,
                                      b = 0, 
                                      l = 0)),
        axis.title.x = element_text(size = 16,
                                    margin = margin(
                                      t = 10, 
                                      r = 0,
                                      b = 0, 
                                      l = 0)),
        title = element_text(size = 16)
  )

g <- function(x){return(0.1/((x + 0.5)^0.25))}
p2 <- data.frame(x = seq(0, 4, by = 0.1)) %>%
  ggplot(aes(x = x)) +
  stat_function(fun = g, col = "blue", linewidth = 1) +
  theme_economist_white(base_size = 24) +
  ggtitle("g(x)") +
  theme(panel.background = element_rect("white"),
        plot.background = element_rect(fill = "white"),
        axis.title.y = element_text(size = 16,
                                    margin = margin(
                                      t = 0, 
                                      r = 10,
                                      b = 0, 
                                      l = 0)),
        axis.title.x = element_text(size = 16,
                                    margin = margin(
                                      t = 10, 
                                      r = 0,
                                      b = 0, 
                                      l = 0)),
        title = element_text(size = 16)
  ) 

grid.arrange(p1, p2, ncol = 2)
```
:::

------------------------------------------------------------------------

::: nonincremental
-   The key to determining linearity is to check the **rate of change**.
:::

-   For a linear function: a one-unit change in `x` corresponds to the same amount of change in `y`, regardless of where the change in `x` occurs.

    -   I.e. as `x` increases from 0 to 1, `y` should increase/decrease by the same amount as if `x` had increased from, say, `2` to `3`.

## Linearity

-   For the function `f` above, this *is* the case:

::: fragment
```{r, fig.height = 3.5}
p1
```
:::

::: {.fragment style="font-size:26px"}
|                |                    |                                    |
|:--------------:|:------------------:|:----------------------------------:|
| `x`: $0 \to 1$ | `y`: $0 \to -1.5$  |  $\Delta$`y`: $(-1.5 - 0) = -1.5$  |
| `x`: $2 \to 3$ | `y`: $-3 \to -4.5$ | $\Delta$`y`: $(-4.5 + 3.5) = -1.5$ |
:::

## Linearity

-   For the function `g` above, this is *not* the case:

::: fragment
```{r, fig.height = 3.5}
p2
```
:::

::: {.fragment style="font-size:26px"}
|                |                       |                                        |
|:---------------------:|:------------------------:|:---------------------:|
| `x`: $0 \to 1$ | `y`: $0.12 \to 0.09$  |  $\Delta$`y`: $(0.09 - 0.12) = -0.03$  |
| `x`: $2 \to 3$ | `y`: $0.08 \to 0.073$ | $\Delta$`y`: $(0.073 - 0.08) = -0.007$ |
:::

## Calculus Aside

-   If you are familiar with calculus, you might note that the above discussion is really just a discussion about derivatives.

-   A function $f()$ is linear if and only if its derivative is constant; otherwise it is nonlinear.

## Linearity in Trends/Associations

-   The notion of in the context of trends is pretty much the same: if a one-unit change in `x` corresponds to the same amount of change in `y` everywhere, then we say that the trend is linear. Otherwise, we say the trend is nonlinear.

-   Also, as mentioned during the workshop, we can extend this language to describe the *association* between two variables.

    -   For example, if the scatterplot of two variables displays a positive linear trend, we might say that the two variables have a positive linear *association*.
